---
title: "Longitudinal model for adolescent cortical change in ABCD"
subtitle: "What the mixed model is, why each term is in it, and what the phenotype means"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    embed-resources: true
    fig-width: 8
    fig-height: 4
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
library(data.table)
library(arrow)
library(lme4)
library(ggplot2)
knitr::opts_chunk$set(dev = "png", dpi = 150)

# Repo root, whether rendered from notebooks/ or the repo root.
ROOT <- if (basename(getwd()) == "notebooks") ".." else "."
source(file.path(ROOT, "R", "model_spec.R"))

RUN <- "thickness_dsk_70_36738d15ec02"   # DK parcellation, release 5.1, genetics config
RUN_DIR <- file.path(ROOT, "out", RUN)
stopifnot(dir.exists(RUN_DIR))
```

## The question this model has to answer

We want the genes that drive adolescent cortical development. That requires a
per-subject *rate of change*, not a per-subject level, and it requires that
rate to come with a known precision — because the next step is a genome-wide
association analysis, and an association test on a phenotype whose measurement
error is unknown cannot be interpreted.

The design constraint is that ABCD is a family-based, multi-site,
irregularly-spaced longitudinal study. In release 5.1 most subjects have two
imaging visits and a minority have three. That is close to the minimum for
estimating an individual slope at all, and it is the single fact that most
shapes everything below.

## Why a mixed model rather than a difference score

The obvious alternative is $(y_2 - y_1)/(t_2 - t_1)$ per subject. It is worth
being explicit about why we do not do that, because the difference score is what
most of the imaging-genetics literature on change has used.

A difference score treats each subject's two measurements as error-free. They
are not: scan-to-scan measurement error in cortical thickness is a substantial
fraction of two years of real change. The difference score therefore has a
variance dominated by noise and — worse — cannot tell you so, because with two
observations and two parameters there is no residual left from which to estimate
error.

A mixed model with random intercept and slope pools information across subjects.
Each subject's slope is shrunk toward the group mean by an amount determined by
how noisy their data are, and that shrinkage factor *is* the reliability of
their slope. This gives three things the difference score cannot: a slope for
subjects with unequal visit counts, a per-subject reliability, and a variance
decomposition separating real between-subject differences in rate from
measurement error.

```{r load-data}
mt <- as.data.table(read_parquet(file.path(RUN_DIR, "model_table.parquet")))

# One region, for exposition. fit_lmm.R runs this per region.
REGION <- "lh_precentral"
d <- mt[label == REGION & !is.na(value)]

cat(sprintf("subjects: %d | observations: %d | families: %d | sites: %d\n",
            uniqueN(d$subject), nrow(d), uniqueN(d$family_id), uniqueN(d$site)))
print(table(visits = d[, .N, by = subject]$N))
```

The visit distribution is the whole problem in one table. Read every result
below in its light.

## The model

```{r formula}
cfg <- list(age_basis = "linear", global_covariate = TRUE,
            re_structure = "slope_correlated",
            site_effect = TRUE, family_effect = FALSE,
            covariates = character(0))
f <- build_formula(cfg)
print(f)
```

Term by term:

**`value ~ age_c`** — `age_c` is age centred on the sample mean, in years. The
fixed `age_c` coefficient is the group-average rate of change. Centring makes
the intercept the fitted value at mean age rather than at age zero, which keeps
the intercept interpretable and stops the intercept–slope correlation from being
an artefact of extrapolating to an age no one was scanned at.

**`(1 + age_c | subject)`** — the scientific target. Two correlated random
effects per subject: their own level and their own rate. The correlation is
estimated rather than assumed zero, because a subject with thicker cortex at
baseline plausibly thins at a different rate, and forcing that covariance to
zero would push it into the residual.

**`(1 | site)`** — 22 sites, essentially constant within subject. As a fixed
effect this costs 21 degrees of freedom and implicitly treats these particular
sites as the population of interest; as a random intercept it costs one variance
parameter and treats them as exchangeable, which is what they are.

**Global mean, split into within- and between-subject parts** — see
@sec-global. This is the term most easily got wrong.

**Family** — omitted here, deliberately. This is the most consequential switch
in the pipeline; see @sec-family.

```{r fit}
fit <- lmer(f, data = d, REML = TRUE, control = lmer_control())
print(summary(fit), correlation = FALSE)
```

## The global-mean covariate, and why it is split in two {#sec-global}

Thickness estimates share a lot of variance across regions: a scan with more
motion produces a thinner-looking cortex everywhere. To study *regional*
development rather than global scaling, we condition on the whole-cortex mean.

But a single global term conflates two different things. A subject whose whole
cortex is thin relative to other subjects is a biological fact about that
subject. A visit at which a subject's cortex measures thin relative to their
*own* other visits is mostly an occasion-level artefact. One `global` term
forces a single coefficient to describe both.

So the global mean enters as two columns: the subject's own mean across visits
(`global_between`) and each visit's deviation from it (`global_within`). This
within/between decomposition matters here specifically because the
within-subject part is where the artefact lives.

```{r global-decomp}
cat(sprintf("SD of within-subject global deviation: %.4f\n", sd(d$global_within)))
cat(sprintf("SD of between-subject global mean:     %.4f\n",
            sd(d[, global_between[1], by = subject]$V1)))
```

The consequence for interpretation: with this covariate in, `age_c` estimates
change in a region *relative to* the cortex-wide average. That is a different
quantity from absolute regional change, and both are legitimate. The pipeline
fits both (`configs/ct_hcp.yaml` vs `configs/ct_hcp_noglobal.yaml`) and they are
not interchangeable — the absolute map answers "where does cortex thin?", the
conditioned map answers "which regions deviate from the global pattern?".

## The family random effect is a phenotype decision, not a fitting detail {#sec-family}

ABCD contains roughly 5,989 families across 6,937 subjects, including about 800
MZ and 1,700 DZ/sibling pairs. The instinct — mine initially — is that a
family-based design demands a family random intercept and that omitting one is
an error.

For estimating fixed effects, that instinct is right. For *extracting a
phenotype for genetic analysis* it is exactly backwards, and the reason is worth
stating carefully.

A random intercept for family absorbs the variance shared between relatives.
Variance shared between relatives is precisely the genetic signal. So
subject-level BLUPs from a model including family are *deviations from the
family mean* — a phenotype from which heritability has been deliberately
removed.

This is not a theoretical worry. Fitting with and without the family term and
computing twin correlations on the whole-cortex mean (207 MZ, 617 DZ/sib pairs):

| phenotype | family RE | MZ *r* | DZ/sib *r* | Falconer *h²* |
|:---|:---|---:|---:|---:|
| intercept | included | 0.14 | −0.62 | — |
| intercept | omitted | 0.88 | 0.47 | 0.83 |
| slope | included | — | — | 0.54 |
| slope | omitted | — | — | 0.58 |

A DZ correlation of −0.62 is biologically impossible. It is the arithmetic
signature of subtracting a family mean from two siblings. Without the family
term the intercept recovers *h²* ≈ 0.83, matching published twin heritability
of cortical thickness.

The slope is far less affected (0.54 vs 0.58) because slope variance is mostly
within-family anyway, but the direction is the same.

**Therefore:** `configs/ct_genetics.yaml` sets `family_effect: false` and is the
config for anything feeding a GWAS. `configs/ct_baseline.yaml` keeps the family
term and is for descriptive fixed-effect inference, where ignoring family
structure would understate standard errors. The two configs produce different
phenotypes on purpose; the run directory records which was used.

## Reading the variance components

```{r varcomp}
vc <- as.data.table(VarCorr(fit))
print(vc[, .(grp, var1, var2, sdcor = round(sdcor, 5))])
```

`sd(age_c | subject)` is the between-subject spread of developmental rate — the
quantity a slope GWAS tries to explain. `sigma` is per-observation measurement
error. Their ratio sets the ceiling on GWAS power.

```{r reliability}
tau_s <- vc[grp == "subject" & var1 == "age_c" & is.na(var2), sdcor]
sigma <- vc[grp == "Residual", sdcor]

# Reliability of an individual slope. Sxx is the within-subject spread of age:
# the lever arm. This is also exactly the mixed model's shrinkage factor.
per <- d[, .(nv = .N, Sxx = sum((age_c - mean(age_c))^2)), by = subject]
per[, rel := tau_s^2 / (tau_s^2 + sigma^2 / pmax(Sxx, 1e-12))]

cat(sprintf("slope reliability: median %.3f (IQR %.3f-%.3f)\n",
            median(per$rel), quantile(per$rel, .25), quantile(per$rel, .75)))
print(per[, .(median_reliability = round(median(rel), 3), n = .N), by = nv][order(nv)])
```

Two things to note. First, this hand calculation reproduces what
`abcd.phenotype` reports for this region (0.033 / 0.118 for 2 / 3 visits),
which is a useful check that the shrinkage in the module is the shrinkage in the
formula and not something else.

Second, `lh_precentral` is a *below-average* region. Across all 68 DK regions
the median slope reliability is 0.12 (2 visits) and 0.16 (3 visits), mean 0.15
and 0.19 — so the numbers printed above are pessimistic relative to the cortex
as a whole, and the good regions reach roughly 0.3. For comparison, the
intercept phenotype has median reliability 0.88 / 0.91. The intercept is a
well-measured trait; the slope, on release 5.1, is not.

This is the headline limitation of release 5.1 and the reason to want 7.0.

Reliability of an individual slope is
$\tau_s^2 / (\tau_s^2 + \sigma^2/S_{xx})$, where $S_{xx}$ is the within-subject
spread of age. Two consequences follow directly, and both are counterintuitive:

1. **Follow-up *span* matters more than visit *count*.** $S_{xx}$ grows
   quadratically with span but only linearly with the number of visits. Three
   visits crammed into two years are worth less than two visits eight years
   apart.
2. **Averaging regions does not rescue it.** The dominant noise is
   per-occasion (motion, positioning) and therefore shared across regions.
   Averaging 68 regions cancels almost none of it — which is why the
   whole-cortex mean slope has *lower* reliability (0.12) than the better
   individual regions.

At reliability $r \approx 0.15$ an association study's observed effect size is
attenuated by $\sqrt{r} \approx 0.39$, so the required sample size for equal
power scales as $1/r \approx 6.7\times$. That is why the slope GWAS on 5.1 is a
pipeline test, not a discovery attempt.

## Individual trajectories, and what shrinkage is doing

```{r trajectories}
#| fig-height: 3.6
set.seed(1)
ids <- sample(unique(d$subject), 60)
bl <- as.data.table(coef(fit)$subject, keep.rownames = "subject")
setnames(bl, "(Intercept)", "b0"); setnames(bl, "age_c", "b1")

pd <- merge(d[subject %in% ids], bl[, .(subject, b0, b1)], by = "subject")
pd[, fitted_line := b0 + b1 * age_c]

ggplot(pd, aes(age_c, value, group = subject)) +
  geom_line(colour = "grey78", linewidth = .3) +
  geom_point(size = .5, colour = "grey55") +
  geom_line(aes(y = fitted_line), colour = "#c0392b", linewidth = .35, alpha = .85) +
  geom_abline(intercept = fixef(fit)[["(Intercept)"]],
              slope = fixef(fit)[["age_c"]], colour = "black", linewidth = .9) +
  labs(x = "age centred on sample mean (years)",
       y = sprintf("thickness, %s (mm)", REGION),
       title = "Grey: observed. Red: shrunken subject fits. Black: group mean.",
       subtitle = "Red lines fan out far less than grey ones — that gap is the measurement error") +
  theme_minimal(base_size = 9)
```

The visual signature of low reliability is that the red subject-specific lines
are far more tightly bunched than the raw grey data. The model is saying that
most of the apparent between-subject variability in rate is noise, and correctly
refusing to propagate it into the phenotype. A difference-score analysis would
have propagated all of it.

## Singular fits are information, not failures

```{r singular}
cat("singular:", isSingular(fit), "\n")
cat("converged:", is.null(fit@optinfo$conv$lme4$messages), "\n")
```

Fitting 68 or 360 regions produces some singular fits — usually a slope
variance at the boundary, or an intercept–slope correlation at ±1. `fit_lmm.R`
records these in `fits/diagnostics.parquet` rather than dropping them, because a
region whose slope variance is zero is making a substantive claim: there is no
detectable between-subject variation in rate there, so nothing for a GWAS to
find. Dropping such regions would bias the regional maps toward regions where
noise happened to resemble signal.

## What runs in production

Everything above is exposition on one region. The pipeline is:

```bash
python -m abcd.assemble configs/ct_genetics.yaml       # long table for a config
Rscript R/fit_lmm.R --run-dir out/<run_id> --cores 8   # fit every region
python -m abcd.phenotype out/<run_id>                  # subject phenotypes + reliability
python -m abcd.gcta_export out/<run_id>                # GCTA/MAGMA inputs
```

Outputs per run: `fits/fixed.parquet` (fixed effects per region),
`fits/varcomp.parquet` (variance components), `fits/diagnostics.parquet`
(convergence, singularity), `phenotypes/phenotypes.parquet` (subject × region ×
{intercept, slope} with reliability).

## Open questions for release 7.0

1. **Does slope reliability actually rise enough?** The $S_{xx}$ expression
   predicts roughly a tripling with four visits over eight years. Check this
   before assuming the GWAS becomes viable — it is a two-line calculation on the
   new visit structure and it determines whether the rest is worth running.
2. **Is linear-in-age still adequate over a longer span?** On 5.1 the spline
   model was not preferred (`R/compare_models.R`), but the span was short. Over
   8+ years the thinning trajectory is likely to curve, and if it does the
   "slope" phenotype must be redefined — a spline coefficient, or an
   area-under-curve summary.
3. **Which metric?** Thickness is the default, but the pipeline is
   metric-agnostic (`metric:` in the config; T1w/T2w ratio already runs). Given
   the analysis above, scan-rescan reliability matters more than effect size
   when choosing.
